---
title: "p8105_hw2_eej2124"
author: "Emma"
date: "2026-10-01"
output: github_document
---

```{r message = F}
library(tidyverse)
library(readxl)
```

# Problem 1 

## Read and clean the CSV data

```{r warning=FALSE}
subway_df = 
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
           na = c("", ".", "NA")) |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(
    entry = case_match(
      entry, 
      c("YES") ~ TRUE,
      c("NO") ~ FALSE 
    )
  )
```
The subway dataset includes `r colnames(subway_df)` variables. These variables are providing information about each NYC MTA subway station, including name, subway line, subway route, and whether the station entrance is ADA compliant. To clean the data, I used the `clean_name()` function from the `janitor` package to convert all variable names to snake case, used the `select()` function to retain only specific variable columns, and finally used the `case_match()` function to convert the character variable `entry` to a logical variable. After cleaning, the dataset has `r ncol(subway_df)` columns and `r nrow(subway_df)` rows. These data are not tidy. 

## Questions

### 1 

```{r}
distinct_stations = 
    subway_df |>
      distinct(line, station_name, ada)

nrow(distinct_stations)
```
There are 465 distinct stations in the MTA subway system. 

### 2 

```{r}
filter(distinct_stations, ada == TRUE) |> 
  nrow()
```
There are 84 ADA-compliant stations in the NYC MTA subway system. 

### 3 

```{r}
filter(subway_df, vending == "NO", entry == "TRUE") |> 
  nrow()
```
69 station entrances or exits without vending allow entrance. 

## Reformatting data 

```{r}
subway_df = 
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
           na = c("", ".", "NA")) |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(
    entry = case_match(
      entry, 
      c("YES") ~ TRUE,
      c("NO") ~ FALSE 
    )
  ) |> 
  mutate(
    route8 = as.character(route8),
    route9 = as.character(route9), 
    route10 = as.character(route10),
    route11 = as.character(route11)
  ) |> 
  pivot_longer(
   route1:route11, 
   names_to = "route_number", 
   values_to = "route_name"
  ) 

subway_df = 
  mutate(drop_na(subway_df, route_name))
```

```{r}
a_train_stations = 
  subway_df |> 
  filter(route_name == "A") |> 
  distinct(line, station_name)
  
nrow(a_train_stations)
```
60 stations serve the A train. 

```{r}
a_train_ada = 
  subway_df |> 
  filter(route_name == "A", ada == TRUE) |> 
  distinct(line, station_name)

nrow(a_train_ada)
```
17 of these stations are ADA compliant. 

# Problem 2 

## Read and clean the Mr. Trash Wheel Excel data 

```{r}
mr_trash_wheel_df = 
  read_excel("data/202610 Trash Wheel Collection Data.xlsx",
             sheet = "Mr. Trash Wheel", 
             range = "A2:N742") |> 
  janitor::clean_names() |> 
  select(dumpster:homes_powered) |> 
  mutate( 
    sports_balls = 
      as.integer(sports_balls), 
    trashwheel = "mr_trash_wheel_df", 
    year = as.numeric(year), 
    plastic_bottles = as.numeric(plastic_bottles)
    )
```
I first imported the Trash Wheel dataset using `read_excel()` from `library(readxl)`. I specified which `sheet()` and `range()` to import so as to exclude the trash wheel figure in the first row and the column sums in the last row. I then cleaned my new data frame using `clean_names()` and used `select()` to remove extra columns. I next used the `mutate()` function to change the `sports_balls` values to whole number integers. I then changed the `year` and `plastic_bottles` variables to numeric, which allowed me to `bind_rows()` later on. Finally, I added an additional column to identify which trash wheel was responsible for which trash collection post-binding. 

## Import, clean, and organize Professor Trash Wheel and Gwynnda data 

```{r warning=FALSE}
prof_trash_wheel_df = 
  read_excel("data/202610 Trash Wheel Collection Data.xlsx",
             sheet = "Professor Trash Wheel", 
             range = "A2:M139") |> 
  janitor::clean_names() |> 
  mutate( 
    trashwheel = "prof_trash_wheel_df", 
    year = as.numeric(year), 
    plastic_bottles = as.numeric(plastic_bottles)
    )
```

```{r}
gwynnda_trash_wheel_df = 
  read_excel("data/202610 Trash Wheel Collection Data.xlsx",
             sheet = "Gwynnda the Good Wheel of the W", 
             range = "A2:L394") |> 
  janitor::clean_names() |> 
  mutate( 
    trashwheel = "gwynnda_trash_wheel_df", 
    year = as.numeric(year), 
    plastic_bottles = as.numeric(plastic_bottles)
    )
```

## Combine all of the trash wheel dataframes  

```{r}
final_trash_wheel_df = 
  bind_rows(mr_trash_wheel_df, prof_trash_wheel_df, gwynnda_trash_wheel_df) |> 
  janitor::clean_names() |> 
  select(-homes_powered) |> 
  relocate(trashwheel) |> 
  pivot_longer(
    plastic_bottles:sports_balls, 
    names_to = "trash_type", 
    values_to = "trash_amount"
  ) 
```
I created my final dataset using `bind_rows()`, which appended all the other data set rows together. In my final Trash Wheel dataset, I removed the `homes_powered` column as it is not dumpster specific. I then tidied the data by moving trash type into its own column and values showing the amount of trash into new column. 

## Trash wheel questions 

```{r}
trash_weight_df = 
  final_trash_wheel_df |> 
  filter(trashwheel == "prof_trash_wheel_df") |> 
  distinct(weight_tons, date) |> 
  pull(weight_tons)

sum(trash_weight_df)
```

```{r}
cig_butt_df = 
  final_trash_wheel_df |> 
  filter(trashwheel == "gwynnda_trash_wheel_df", month == "June", year == 2022, trash_type == "cigarette_butts") |> 
  pull(trash_amount)

sum(cig_butt_df)
```
The final dataset includes information on `r colnames(final_trash_wheel_df)`. The total weight of the trash collected by Professor Trash Wheel was `r sum(trash_weight_df)`. The total number of cigarette butts collected by Gwynnda in June 2022 was `r sum(cig_butt_df)`.

# Problem 3 

## Import and clean baseline demographics data 

```{r}
baseline_df = 
  read_csv("data/data_mci/MCI_baseline.csv", 
           skip = 1, 
           na = c("", ".", "NA")) |> 
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex, 
      1 ~ "male",
      0 ~ "female"
      ), 
    apoe4 = case_match(
      apoe4, 
      1 ~ "apoe4_carrier", 
      0 ~ "apoe4_non_carrier"
    )
  )  |> 
  filter(age_at_onset > current_age|is.na(age_at_onset))
```
I imported the baseline data using `read_csv()`, skipped the first row, and used `clean_names()`. In order to change the `sex` and `apoe4` encoding from numeric to character, I used `mutate()` and `case_match()` functions. To remove participants who do not meet the inclusion criteria (i.e. remove participants who had MCI at baseline), I used the `filter()` function to keep participants who have an age of onset older than the current age (the age at baseline) OR participants who did not develop MCI at follow ups. This allows us to exclude participants who had MCI at the beginning of the study.

## Baseline questions

```{r}
nrow(filter(baseline_df, age_at_onset != "NA"))
```

```{r}
mean(pull(baseline_df, current_age))
```

```{r}
nrow(filter(baseline_df, sex == "female"))
nrow(filter(baseline_df, apoe4 == "apoe4_carrier", sex == "female"))
```

`r nrow(baseline_df)` participants were recruited to this study. Of those, `r nrow(filter(baseline_df, age_at_onset != "NA"))` participants developed MCI. The average baseline age was `r round(mean(pull(baseline_df, current_age)), 2)`. Of the women in the study, `r nrow(filter(baseline_df, apoe4 == "apoe4_carrier", sex == "female"))` out of `r nrow(filter(baseline_df, sex == "female"))` women were APOE4 carriers. 

## Import and clean biomarker values data

```{r}
biomarkers_df = 
  read_csv("data/data_mci/mci_amyloid.csv", 
           skip = 1, 
           na = c("", ".", "NA")) |> 
  janitor::clean_names() |> 
  pivot_longer(
    baseline:time_8,
    names_to = "years_elapsed", 
    values_to = "amyloid_42_40_ratio"
  ) |> 
   mutate(
    years_elapsed = case_match(
      years_elapsed, 
     "baseline" ~ 0, 
     "time_2" ~ 2,
     "time_4" ~ 4, 
     "time_6" ~ 6,
     "time_8" ~8
    )
  ) |> 
  rename(id = study_id)
```

I carried out a similar process of importing and cleaning for the `biomarkers_df` dataset. 
In order to tidy the data, I used `pivot longer()` because important information, specifically the years elapsed since baseline, was stored in the column names, and biomarker values were stored in the data cells. I added a column called `years_elapsed` to detail the number of years elapsed since the baseline, and I moved the biomarker values into a new column called `amyloid_42_40_ratio`. Finally, I mutated the data in the `years_elapsed` column to numerical values of years that had passed using `case_match()`. 

```{r}
non_overlap_participants = 
  anti_join(baseline_df, biomarkers_df, by = "id")
```

```{r}
non_overlap_participants_1 = 
  anti_join(biomarkers_df, baseline_df, by = "id") |> 
  distinct(id)
```
`r nrow(non_overlap_participants)` participants appear only in `baseline_df` and not in `biomarkers_df`, and `r nrow(non_overlap_participants_1)` participants appear only in `biomarkers_df` and not in `baseline_df`.   

# Combine the data 
```{r}
mci_study_df = 
  inner_join(baseline_df, biomarkers_df)
```
The final dataset has `r nrow(mci_study_df)` rows and `r ncol(mci_study_df)` columns. It describes the `r colnames(mci_study_df)` of the participants. 

# Export final CSV
```{r}
write_csv(mci_study_df, "mci_study_df.csv")
```





