---
title: "p8105_hw2_eej2124"
author: "Emma"
date: "2026-10-01"
output: github_document
---

```{r message = F}
library(tidyverse)
library(readxl)
```

# Problem 1 

## Read and clean the csv data
Read the data, clean the names, and retain line, station name, station latitude / longitude, routes served, entrance type, entry, vending, and ADA compliance. Converting entry variable from character to logical. 
```{r}
subway_df = 
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", na = c("", ".", "NA")) |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(
    entry = case_match(
      entry, 
      c("YES") ~ TRUE,
      c("NO") ~ FALSE 
    )
  )
```
The subway dataset includes `r colnames(subway_df)` variables. These variables are providing information about each NYC MTA subway station, including name, subway line, subway route, and whether the station entrance is ADA compliant. To clean the data, I used the `clean_name` function from the `janitor` package to convert all variable names to snake case, used the `select` function to retain only specific variable columns, and finally used the `case_match` function to convert the character variable `entry` to a logical variable. After cleaning, the dataset has `r ncol(subway_df)` columns and `r nrow(subway_df)` rows. These data are tidy. 

# are these data tidy?

## Questions
### 1 

### 2 

### 3 

## Reformatting data 

# Problem 2 

## Read and clean the Mr. Trash Wheel Excel data 

```{r}
trash_wheel_df = 
  read_excel("data/202509 Trash Wheel Collection Data.xlsx",
             sheet = "Mr. Trash Wheel") |> 
  janitor::clean_names()
```
# Omit non data entries and rows that do not include dumpster-specific data, round number of sports balls using round() and convert to integer using as.integer 

## Import, clean, and organize Professor Trash Wheel and Gwynnda data 
```{r}
prof_trash_wheel_df = 
  read_excel("data/202509 Trash Wheel Collection Data.xlsx",
             sheet = "Professor Trash Wheel") |> 
  janitor::clean_names()
```

```{r}
gwynnda_trash_wheel_df = 
  read_excel("data/202509 Trash Wheel Collection Data.xlsx",
             sheet = "Gwynns Falls Trash Wheel") |> 
  janitor::clean_names()
```

# Combine with Mr. Trash Wheel, do other stuff 

# Problem 3 

## Import and clean baseline demographics data 
```{r}
baseline_df = 
  read_csv("data/data_mci/MCI_baseline.csv", na = c("", ".", "NA")) |> 
  janitor::clean_names()
```

## Import and clean biomarker values data
```{r}
biomarkers_df = 
  read_csv("data/data_mci/mci_amyloid.csv", na = c("", ".", "NA")) |> 
  janitor::clean_names()
```

